---
title: "Memo - Determing the cutoff of positive and negative peaks (fantom fetal brain cage peak from Zhongshan 06/01/18)"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 3
date: "`r format(Sys.time(), '%B %d, %Y')`"
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=2000)
library(pander)
```

# About data

Obtained from `hg19.cage_peak_phase1and2combined_tpm.osc.txt.gz` - CAGE peak based expression table (RLE normalized) for human samples [see edgeR documentation for further details about normalization method]

Only fetal brain is extracted (Zhongshan's part)

# Preprocessing

Obtained data from `/project2/xinhe/zhongshan/noncodingRNA/FANTOM/hg19.cage_peak_phase1and2combined_tpm.osc.txt.ordered.CNhs11797`

## Negative sequences

All peaks with zero entry are defined as negative sequences.

```{bash, eval=FALSE}
$ cat fantomFetalBrainCage.bed |awk -F"\t" -v OFS="\t" '{if($5==0) print $1,$2,$3,$5}' > fantomFetalBrainCage.negative.bed
```

## Pre-positive sequences

Obtain the 'pre-positive' sequences by extracting all sequences with non-zero entry.

```{bash, eval=FALSE}
$ cat fantomFetalBrainCage.bed |sort -k5 -nr|awk -F"\t" '{if($5!=0) print $5}' > fantomFetalBrainCage.pre-positive.bed
```

## Distribution of pre-positive peaks

```{r}
prepos <- read.table('../data/fantomFetalBrainCage.pre-positive.bed')
hist(log10(prepos$V1), xlab = 'log10(score)', main = '')
Fn <- ecdf(log10(prepos$V1))
plot(Fn)
cutoff.log <- c(0, 1, 2)
df <- data.frame(cutoff = 10^cutoff.log, topNprecent = 1 - Fn(cutoff.log))
pander(df)
```

# Definition used downstream

```{bash, eval=FALSE}
$ cat fantomFetalBrainCage.bed |sort -k5 -nr|awk -F"\t" -v OFS="\t" '{if($5>10) print $1,$2,$3,$5}' > fantomFetalBrainCage.gt10.bed
$ cat fantomFetalBrainCage.bed |awk -F"\t" -v OFS="\t" '{if($5==0) print $1,$2,$3,$5}' > fantomFetalBrainCage.eq0.bed
```
