---
title: "Deep brain - Motif Analysis - Sliding Window Mask Size = 22bp"
author: Yanyu Liang
date: "`r format(Sys.time(), '%d %B, %Y')`"
---
  
<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=2000)
```

# Procedure

[AME](http://meme-suite.org/tools/ame) (Analysis of Motif Enrichment; tool in MEME Suite) is used to do motif enrichment analysis. For a given motif set and a set of sequences, test whether a motif occurs significantly in the sequences. 

Some windows were extracted and fed to AME to perform motif enrichment analysis (search in JASPAR 2016 vertebrates). Here top 0.1% windows that decrease the score most in positive sequences and top 0.1% windows that increase the score most in negative sequences and randomly selected windows in positive sequences were analyzed respectively.

Results of randomly selected windows can be found:

  - [E081 top 0.1% positive](AME_results_E081.top0.1_positive.html), 
  - [E081 random positive](AME_results_E081.random_positive.html),
  - [E081 top 0.1% negative](AME_results_E081.top0.1_negative.html),
  - [Noonan top 0.1% positive](AME_results_Noonan.top0.1_positive.html), 
  - [Noonan random positive](AME_results_Noonan.random_positive.html),
  - [Noonan top 0.1% negative](AME_results_Noonan.top0.1_negative.html),
  - [hNSC-50 top 0.1% positive](AME_results_hNSC-50.top0.1_positive.html), 
  - [hNSC-50 random positive](AME_results_hNSC-50.random_positive.html),
  - [hNSC-50 top 0.1% negative](AME_results_hNSC-50.top0.1_negative.html),

# GO term analysis of hitted TF

Here JASPAR TFs is selected as the background protein list.

## Overview

```{r, echo=FALSE, message=FALSE}
clean_gene_name <- function(x){
  e <- strsplit(x, '::')
  re <- c()
  for(i in e){
    i <- strsplit(i, '(', fixed = T)[[1]]
    re <- c(re, i[1])
  }
  return(re)
}
go_term_bp <- function(gene_cleaned, back_cleaned){
  ensembl = useMart(biomart = "ENSEMBL_MART_ENSEMBL",dataset="hsapiens_gene_ensembl", host = "jul2015.archive.ensembl.org")
  goids = getBM(attributes = c('hgnc_symbol', 'go_id'), 
                filters = 'hgnc_symbol', 
                values = back_cleaned, 
                mart = ensembl)
  gene2GO <- lapply(split(goids, goids['hgnc_symbol']), function(x){x[, 2]})
  gene_list_analysis <- factor(as.integer(back_cleaned %in% gene_cleaned))
  names(gene_list_analysis) <- back_cleaned
  myGOdata <- new("topGOdata", description="Important Sequences in Sliding Window Analysis Mapped to TF Genes", ontology="BP", allGenes=gene_list_analysis,  annot = annFUN.gene2GO, gene2GO = gene2GO)
  resultFisher <- runTest(myGOdata, algorithm="classic", statistic="fisher") 
  allRes <- GenTable(myGOdata, classicFisher = resultFisher, orderBy = "resultFisher", ranksOf = "classicFisher", topNodes = 50)
  return(allRes)
  
}
read_and_clean <- function(filename){
  temp <- tryCatch({
    read.table(filename, sep = ' ', skip = 12)
  },
    error=function(cond) {
      return(NULL)
  })
  if(!is.null(temp)){
    temp.gene <- temp$V7
    temp.gene_cleaned <- apply(as.matrix(temp.gene, ncol=1), 1, clean_gene_name)
    return(temp.gene_cleaned)
  }else{
    return(temp)
  }
}

background <- read.table('../data/background_jaspar_gene.txt')
background.cleaned <- apply(background, 1, clean_gene_name)

datasets <- c('E081', 'Noonan', 'hNSC-50')
conditions <- c('top0.1_positive', 'top0.1_negative', 'random_positive')
mydata <- list()
tf.counts <- c()
for(dataset in datasets){
  tf.count <- c()
  for(cond in conditions){
    filestring <- paste(dataset, cond, sep = '.')
    gene.list <- read_and_clean(paste0('../data/ame_', filestring, '.txt'))
    if(is.null(gene.list)){
      tf.count <- c(tf.count, 0)
    }else{
      tf.count <- c(tf.count, length(gene.list))
      mydata[[filestring]] <- gene.list
    }
  }
  tf.counts <- rbind(tf.counts, tf.count)
}
tf.counts <- as.data.frame(tf.counts)
row.names(tf.counts) <- datasets
colnames(tf.counts) <- conditions
```

```{r, results='asis'}
library(pander)
pandoc.table(tf.counts)
```

```{r, echo=FALSE, message=FALSE}
library(topGO)
library(biomaRt)

go.results <- list()

for(dataset in datasets){
  for(cond in conditions){
    filestring <- paste(dataset, cond, sep = '.')
    if(!is.null(mydata[[filestring]])){
      go.results[[filestring]] <- go_term_bp(mydata[[filestring]], background.cleaned)
    }
  }
}
```

```{r, results='asis'}
panderOptions('knitr.auto.asis', FALSE)
panderOptions('table.split.table', Inf)
for (i in names(go.results)){
  cat('\n')
  cat("## ", i, "\n")
  pander(go.results[[i]])
  cat('\n')
}
```

## Summary

The top 1% of the positive windows in E081 and Noonan hit in enriched in neuron development process.
